D3064R0: Bringing the Laws of Physics, Computer-System Constraints, and Compilers to Bear on the OOTA Problem

### 2024-02-16 cuttingroom.tex
Move "complex sdep" to cbmc portion of cutting-room.tex
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

But first, Appendix~\ref{app:Informal Definition of Semantic Dependency}
outlines an informal informal definition of ``semantic dependency''
that can be used in \co{cbmc} test cases.

@@

>>>>

But first, Appendix~\ref{app:Informal Definition of Semantic Dependency}
outlines an informal informal definition of ``semantic dependency''
that can be used in \co{cbmc} test cases, and then
Appendix~\ref{sec:Executions Can Be Complex} gives some examples
of mathematically complex semantic dependencies.

@@

\subsection{Executions Can Be Complex}
\label{sec:Executions Can Be Complex}

Consider this example:

\begin{quote}
\begin{verbatim}
x = horribly_complex(y, z);
\end{verbatim}
\end{quote}

Here, the presence or absence of a semantic dependency from \co{y} to
\co{x} depends not only on the value of \co{z}, but also and execution
that passes through the code in a function that, although single threaded,
is horribly complex.

A special case is as follows:

\begin{quote}
\begin{verbatim}
x = y & 0x1 ? do_odd(y) : do_even(y);
\end{verbatim}
\end{quote}

The key point here is that different values might result in almost
completely disjoint executions.

Fortunately, there are now tools that can help.
For example, the SAT-solver-based \co{cbmc} tool
has been used to mechanically verify signicant portions of Linux-kernel
RCU from the C-language source
code~\cite{LihaoLiang2016VerifyTreeRCU,LanceRoy2017CBMC-SRCU}.
Another tool, Nidhugg, which is based on partial-order
reduction, has been used to carry out a similar
verification~\cite{MichalisKokologiannakis2017NidhuggRCU,SMC-TreeRCU,MichalisKokologiannakis2019RCUstatelessModelCheck}.
Both tools can easily check whether or not a given dependency is semantic
for at least some approximate definitions of ``semantic dependency''.
% An example use of \co{cbmc} is illustrated in
% Appendix~\ref{sec:Evaluating sdep Using cbmc}.

There are also many other tools, but failing that, it is always possible
to fall back to older software-verification tools such as manual
code inspection.

But consider the following code fragment:

\begin{quote}
\begin{verbatim}
x = collatz_cycle_min(y);
\end{verbatim}
\end{quote}

Here, \co{collatz_cycle_min()}\footnote{
	\url{https://en.wikipedia.org/wiki/Collatz_conjecture}}
takes its arbitrary-precision-integer argument and repeatedly either
divides it by two if it is even, or multiplies it by three and adds one
if it is odd.
Should the resulting sequence of arbitrary-precision integers reach
a cycle, the function returns the minimum element of that cycle.
Leaving aside the practical complication of a value of \co{y} that
consumes more than half of the available memory, there is reason to
believe that the above code fragment will always assign the value 1 to
\co{x}.
Except that this reason for belief is based on the Collatz conjecture,
which is currently unproven and might never be proven (or disproven).

A similar example could be constructed from Fermat's Last Theorem
(FLT),\footnote{
\url{https://en.wikipedia.org/wiki/Fermat\%27s_Last_Theorem}}
which was first stated in the year 1637 and finally proved more than
300~years later, in 1994.
Therefore, it was only after 1994 that certain types of executions
relying on FLT were known not to have a semantic dependency.

Many analysis tools would respond to these examples by failing, for example,
by consuming all available CPU and memory.

One could quite reasonably argue that these two examples are irrelevant
to any real-world C++ implementation.
After all, evidence to date suggests that code relying on either the
Collatz Conjecture or FLT is quite rare, and thus not worth optimizing.
And the value of any given optimization is another real-world constraint
on the problem of locating OOTA cycles, but one that will not be considered
further in this document.

\emph{Because of unsolved mathematical problems, precise definitions of
``execution'' have limitations.
Therefore, C++ implementations must assume that a dependency passing
through a given execution is a semantic dependency unless and until it
can prove otherwise.}
